TrueSkill Through Time: reliable initial skill estimates and historical comparability with \proglang{Julia}, \proglang{Python}, and \proglang{R}

### 2023-05-18 doc/article.tex
Provisonally accepted

\newif\ifen
\newif\ifes
\newcommand{\en}[1]{\ifen#1\fi}
\newcommand{\es}[1]{\ifes#1\fi}
\entrue
\usepackage[utf8]{inputenc}
\usepackage{paracol}
\usepackage{multirow}

%% recommended packages orcidlink,
\usepackage{orcidlink,thumbpdf,lmodern}

@@

% by: https://tex.stackexchange.com/questions/74636/mla-package-and-thumbpdf
\makeatletter
\@namedef{ver@thumbpdf.sty}{}
\makeatother

@@

% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={black!50!black},
%     citecolor={black!50!black},
%     urlcolor={black!80!black}
% }

@@

%% - \Abstract{} almost as usual
\Abstract{
\en{Humans develop complex skills through time. }%
\es{Los humanos desarrollamos habilidades complejas a través del tiempo. }%
\en{Knowing how individual abilities change is essential in a wide range of activities. }%
\es{Saber c\'omo cambian las capacidades individuales es esencial en varias actividades. }%
\en{The most commonly used skill estimators in industry and academia (such as Elo, Glicko, IRT, and TrueSkill) cannot obtain reliable initial estimates or guarantee comparability between estimates distant in time and space. }%because they do not propagate historical information properly through the system. }%
\es{Los estimadores de habilidad más utilizados en la industria y la academia (como Elo, Glicko, IRT y TrueSkill) no pueden obtener estimaciones iniciales fiables ni garantizar la comparabilidad entre estimaciones distantes en el tiempo y el espacio. }%debido a que no propagan la informaci\'on hist\'orica apropiadamente a través del sistema. }%
\en{The model TrueSkill Through Time (TTT) propagates all historical information throughout a single causal network, providing estimates with low uncertainty at any given time, enabling reliable initial skill estimates, and ensuring historical comparability. }%
\es{El modelo TrueSkill Through Time (TTT), propagando toda la informaci\'on hist\'orica a través de una única red causal, proporciona estimaciones con baja incertidumbre en todo momento, permitiendo estimaciones iniciales de habilidad fiables y asegurando su comparabilidad hist\'orica. }%
\en{The improvement is evident since their prior predictions are several orders of magnitude more accurate than those obtained with other models. }%
\es{La mejora es evidente en tanto sus predicciones a priori son varios \'ordenes de magnitud más precisas que las obtenidas con los modelos mencionados. }%
\en{In addition to being intuitive for users, the TTT model achieves similar or even better results than more complex ones, such as KickScore. }%
\es{Además de ser intuitivo para los usuarios, el modelo TTT logra resultados similares y hasta mejores que modelos más complejos, como KickScore. }%
\en{Although the TTT model was published more than a decade ago, it was not available until now in the programming languages with the largest communities. }%
\es{Si bien el modelo TTT fue publicado hace más una década, no se encontraba disponible hasta ahora en los lenguajes de programaci\'on con mayores comunidades. }%
\en{Here we offer the first software for \proglang{Julia}, \proglang{Python}, and \proglang{R}, accompanied by a detailed overview for the general public and an in-depth scientific explanation. }%
\es{Aquí ofrecemos el primer software para \proglang{Julia}, \proglang{Python} y \proglang{R}, acompañado por una descripci\'on detallada para el p\'ublico general, y una explicaci\'on cient\'ifica en profundidad. }%
%%
\en{After illustrating its basic mode of use, we show how to estimate the learning curves of historical players of the Association of Tennis Professionals. }%
\es{Luego de ilustrar su modo de uso básico, mostramos c\'omo estimar la curvas de aprendizaje de los jugadores hist\'oricas de la Asociaci\'on de Tenistas Profesionales. }%
\en{Analytical approximation methods and message-passing algorithms allow inference to be solved efficiently using any low-end computer, even in causal networks with millions of nodes and irregular structures. }%
\es{Los métodos de aproximación analítica y los algoritmos de pasaje de mensajes permite resolver la inferencia de forma eficiente utilizando cualquier ordenador de gama baja, incluso en redes causales con millones de nodos y estructuras irregulares. }%
} 
\Keywords{%
\en{Learning, skill, Bayesian inference, gaming, education, sports, \proglang{Julia}, \proglang{Python},  \proglang{R}}%
\es{Aprendizaje, habilidad, bayes, videojuego, educaci\'on, deporte, \proglang{Julia}, \proglang{Python},  \proglang{R}}%
}
\Plainkeywords{%
\en{Learning, skill, bayesian inference, gaming, education, sports, Julia, Python, R}%
\es{Aprendizaje, habilidad, bayes, videojuego, educaci\'on, deporte, Julia, Python, R}%
}

@@

\section{\en{Introduction}\es{Introducci\'on}} \label{sec:intro}

\en{Humans develop and acquire complex skills because of an special integration of biological, cognitive and social processes~\citep{Koster2020}. }%
\es{Los humanos desarrollamos habilidades complejas gracias a una integraci\'on especial de los procesos biol\'ogicos, cognitivos y sociales~\citep{Koster2020}. }%
%
\en{Our exceptional cognitive ability to imitate, combined with long periods of juvenile dependency and postreproductive life span, allows humans to learn things from others and transmit innovations through generations~\citep{Richerson2020}. }%
\es{Nuestra extraordinaria capacidad para imitar, combinada con los largos per\'iodos de aprendizaje juvenil y vida posreproductiva, permite a los humanos aprender de los dem\'as y transmitir las innovaci\'on a trav\'es de la generaciones~\citep{Richerson2020}. }%
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{Derex2020}. }%
\es{Al ser un proceso poblacional, la adaptaci\'on humana tambi\'en se ve afectada por caracter\'isticas demogr\'aficas, como el tama\~no y estructura de las poblaciones~\citep{Derex2020}. }%

\en{Knowing how individual skills change over time is essential in the educational system and in the labor market. }%
\es{Conocer c\'omo cambian las habilidades individuales a lo largo del tiempo es esencial en el sistema educativo y laboral. }%
%
\en{Since skills are hidden variables, the best we can do is estimating them based on its direct observable consequences: the outcome of problem-solving and competitions. }%
\es{Dado que son variables ocultas, lo mejor que podemos hacer es estimarlas a partir de sus consecuencias observables directas: el producto de la resoluci\'on de problemas y competencias. }%
%
\en{However, estimating learning curves is a sensitive issue, especially when they are intended to be used to make decisions that may impact individuals. }%
\es{Sin embargo la estimaci\'on de curvas de aprendiaje es tema sensible, especialmente cuando se las pretende utilizar para tomar decisiones que pueden impactar a las personas. }%
%
\en{Considering only the frequency of positive results as an indicator of the individuals' ability could lead to wrong approximations, mainly because the outcome also depends on the difficulty of the challenge. }%
\es{Considerar s\'olo la frecuencia de resultados positivos como indicador de la habilidad de los individuos puede conducir a aproximaciones erroneas, fundamentalmente porque su valor depende tambi\'en de la dificultad de los desaf\'ios. }%
%
\en{For this reason, all widely used skill estimators use pairwise comparisons. }%
\es{Por esta raz\'on, todos los estimadores de habilidad ampliamente usados se basan en comparaciones por pares. }%
%
\en{Since the first generative models, proposed almost a century ago by~\cite{Thurstone1927} and~\cite{Zermelo1929}, it is assumed that the probability of the observed result $r$ depends on the performance $p$ of an agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$. }%
\es{Desde los primeros modelos generativos, propuestos hace casi un siglo por~\cite{Thurstone1927} y~\cite{Zermelo1929}, se supone que la probabilidad de un resultado observado $r$ depende del rendimiento $p$ del agente $i$ y de su oponente $j$, expresada como $P(\, r \,|\, p_i, \, p_j \,)$. }%
%
\en{The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, leading to a breakthrough that took place when~\cite{Elo2008} developed a methodology for the US Chess Federation (USCF), which is used by the International Chess Federation (FIDE) nowadays. }%
\es{El campo sigui\'o progresando con los trabajos de \cite{Bradley1952} y~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, que condujeron al gran avance que tuvo lugar cuando~\cite{Elo2008} desarroll\'o una metodología para la Federaci\'on de Ajedrez de los Estados Unidos (USCF), adoptada hasta el d\'ia de hoy por la Federaci\'on Internacional de Ajedrez (FIDE). }%

\en{All currently used skill estimators share some variant of the probabilistic model proposed by Elo. }%
\es{Todos los estimadores de habilidad ampliamente utilizados en la actualidad comparten alguna variante del modelo probabil\'istico propuesto por Elo~\citep{Glickman1999, Herbrich2007, VanDerLinden2016}. }%
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

    \node[latent, above=of r, yshift=-0.45cm] (d) {$d$} ; %
    \node[const, right=of d] (dd) {\normalsize $ d = p_i-p_j$}; 
    \node[const, left=of d, xshift=-1.35cm] (d_name) {\small \en{Difference}\es{Diferencia}:};

    \node[latent, above=of d, xshift=-0.8cm, yshift=-0.45cm] (p1) {$p_i$} ; %
    \node[latent, above=of d, xshift=0.8cm, yshift=-0.45cm] (p2) {$p_j$} ; %
    \node[const, left=of p1, xshift=-0.55cm] (p_name) {\small \en{Performance}\es{Desempeño}:};

@@

    \node[const, left=of s1, xshift=-.85cm] (s_name) {\small \en{Skill}\es{Habilidad}:};

@@

}
     \caption{
     \en{Generative model in which skills cause the observable results mediated by the difference of hidden performances, $d =p_i - p_j$, both random variables around their unknown true skill, $p \sim \N(s,\beta^2)$. }%
    \es{Modelo generativo en el que las habilidades causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos, $d=p_i-p_j$, ambas variables aleatorias centradas en la verdadera habilidad, $p \sim \N(s,\beta^2)$. }%
    %
    \en{The one with the highest performance wins, $r = (d > 0)$. }%
    \es{Quien haya obtenido mayor rendimiento gana, $r = (d > 0)$. }%
    %
    \en{Observable variables are painted gray, hidden in white, and constants are shown as black dots. }%
    \es{Las variables observables se pintan de gris, las ocultas en blanco, y las constantes se muestran como puntos negros. }%
    }
    \label{fig:generative_model}
\end{figure}
%
\en{Figure~\ref{fig:generative_model} provides a causal model in which skills generate the observable result. }%
\es{La figura~\ref{fig:generative_model} ofrece un modelo causal en el que las habilidades generan el resultado observable. }%
%
\en{The agents exhibit different performances at each event, varying around their true skill, $\N(p\,|\,s,\beta^2)$. }%
\es{Los agentes exhiben distintos desempe\~nos en cada evento, que var\'ian alrededor de su verdadera habilidad, $\N(p\,|\,s,\beta^2)$. }%
%
\en{The model assumes that the agent with the highest performance wins, $r = (p_i > p_j)$. }%
\es{El modelo supone que gana el agente con mayor rendimiento, $r = (p_i > p_j)$. }%
%
\en{The parameter $\beta^2$, being the same for all agents, acts as the scale of the estimates: skills at a distance of one $\beta$ mean a \SI{76}{\percent} probability of winning, independent of the absolute value of the estimates. }%
\es{El parámetro $\beta^2$, al ser el mismo para todos los agentes, act\'ua como la escala de las estimaciones: habilidades a una distancia de un $\beta$ implica 76\% probabilidad de ganar, independiente del valor absoluto de las estimaciones }%

\en{Skill estimators commonly used in the video game industry and academia, such as Elo, Glicko~\citep{Glickman1999}, TrueSkill~\citep{Herbrich2007}, and IRT~\citep{VanDerLinden2016}, cannot obtain reliable initial estimates nor ensure historical comparability because they propagate information in only one direction through the system, using the last posterior as the prior probability to the next event (\emph{filtering} approach). }%
\es{Los estimadores de habilidad comúnmente utilizados en la industria del video juego y la academia, como Elo, Glicko~\citep{Glickman1999}, TrueSkill \citep{Herbrich2007} e IRT \citep{Fox2010,VanDerLinden2016}, no pueden obtener estimaciones iniciales fiables ni asegurar la comparabilidad histórica debido a que propagan la información en una sola dirección a través del sistema, utilizando el último posterior como prior del siguiente evento (enfoque \emph{filtering}). }%
%
\en{In contrast, TrueSkill Through Time (TTT) model~\citep{Dangauthier2007} propagates all historical information throughout the entire network of events, providing estimates with low uncertainty at any given time, offering reliable initial skill estimates, and guaranteeing comparability between estimates which are distant in time and space (\emph{smoothing} approach). }%
\es{En cambio, el modelo TrueSkill Through Time (TTT)~\citep{Dangauthier2007} propaga la información histórica a través de toda la red, proporcionando estimaciones con baja incertidumbre en cualquier momento, ofreciendo estimaciones iniciales de habilidad fiables y garantizando la comparabilidad de las estimaciones distantes en el tiempo y el espacio (enfoque \emph{smoothing}). }%
%
\en{\cite{Coulom2008} (Whole History Rating) and \cite{Maystre2019} (KickScore) implemented similar approaches based on Laplacian approximations and Gaussian processes, respectively. }%
\es{Enfoque similares fueron implementados por \cite{Coulom2008} (Whole History Rating) y \cite{Maystre2019} (KickScore), basados en aproximaciones laplacianas y procesos gaussianos respectivamente. }%

\en{In Table~\ref{Tab:Models}, we compare our implementation of the TTT algorithm against KickScore, TrueSkill, Elo, and a ``Constant'' model in which skills do not change over time. }%
\es{En la tabla~\ref{Tab:Models} comparamos nuestra implementaci\'on del algoritmo TTT contra KickScore TrueSkill, Elo, y un modelo ``Constante'' en el que las habilidades no cambian en el tiempo. }%
%
\en{To evaluate the models, we must calculate their probability given the data, }%
\es{Para evaluar los modelos debemos calcular su probabilidad dados los datos, }%
%
\begin{equation}
 P(\text{Model\es{o}}|\text{Dat\en{a}\es{os}}) = \frac{P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}})P(\text{Model\es{o}})}{P(\text{Dat\en{a}\es{os}})}
\end{equation}
%
\en{When we do not have access to all models, we cannot compute the probability $P(\text{Model}|\text{Data})$, but we can compare models. }%
\es{Cuando no tenemos acceso a todos los modelos, no podremos calcular la probabilidad $P(\text{Modelo}|\text{Datos})$, pero sí podremos comparar modelos. }%
%
\begin{equation}\label{eq:bayes_factor}
 \frac{P(\text{Model\es{o}}_i|\text{Dat\en{a}\es{os}})}{P(\text{Model\es{o}}_j|\text{Dat\en{a}\es{os}})} = \frac{P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}_i)P(\text{Model\es{o}}_i)}{P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}_j)P(\text{Model\es{o}}_j)} \overset{\text{\en{if}\es{si} } *}{=} \frac{P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}_i)}{P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}_j)} 
\end{equation}
%
\en{This expression is known as the \emph{Bayes factor}, usually expressed using a logarithmic scale to report the difference in orders of magnitude. }%
\es{Esta expresi\'on se conoce como el \emph{Bayes factor}, el cual se suele expresar en escala logar\'itmica para reportar la diferencia en \'ordenes de magnitud. }%
%
\en{When we have no prior preference over any model (\en{if}\es{si} $*$), we only need to compare the prior predictions of the models. }%
\es{En el caso especial en el que no tenemos preferencia a priori sobre ning\'un modelo (\en{if}\es{si} $*$), s\'olo debemos comparar las predicciones a priori de los modelos. }%
%
\begin{equation}
P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}) = P(d_1|\text{Model\es{o}})P(d_2|d_1,\text{Model\es{o}}) \dots P(d_n|d_1,\dots, d_{n-1},\text{Model\es{o}})
\end{equation}
%
\en{where each element of the product is a prediction of the following data point based on the previous data set and the model. }%
\es{donde cada elementos de la productoria es una predicci\'on del siguiente dato en base a los datos previos y el modelo. }%
%
\en{In this work, we report the geometric mean because it can be interpreted as the ``characteristic'' prediction of the model, }%
\es{En este trabajo también vamos a reportare la media geométrica debido a que puede interpretarse como la predicci\'on ``característica'' del modelo, }%
%
\begin{equation}\label{eq:gmean}
P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}) = \text{geometric mean}(P(\text{Dat\en{a}\es{os}}|\text{Model\es{o}}))^{|\text{Dat\en{a}\es{os}}|}
\end{equation}
%
\en{where $|\text{Data}|$ represents the size of the data set. }%
\es{donde $|\text{Datos}|$ representa el tamaño del conjunto de datos. }%

\en{We follow the methodology employed by \cite{Maystre2019}, analyzing the four databases on which they reported their prior predictions: ATP (tennis), NBA (basketball), FIFA (football), and FIDE (chess). }%
\es{Para realizar la comparaci\'on seguiremos la metodología empleada por \cite{Maystre2019} analizando las cuatro base de datos sobre las cuales ellos reportan sus predicciones a priori: ATP (tenis), NBA (baloncesto), FIFA (f\'utbol), y FIDE (ajedrez). }%
\en{This methodology consists of splitting the data into two subsets: the initial \SI{70}{\percent} used to train the hyperparameters and the last \SI{30}{\percent} to evaluate the models. }%
\es{Usamos el \SI{70}{\percent} inicial de la base de datos para entrenar los hiperparámetros y el \SI{30}{\percent} final para evaluar los modelos. }%
\en{To predict the outcome of one observation at time $t$, we use all the data up to the day preceding $t$ (in both training and test sets). }%
\es{Para predecir el resultado de una observaci\'on en el tiempo $t$, ellos utilizan todos los datos (en los conjuntos de entrenamiento y de prueba) hasta el día anterior a $t$. }%
\en{Table~\ref{Tab:Models} summarizes the model comparisons in each of the four databases. }%
\es{La tabla~\ref{Tab:Models} se resumen las comparaciones de los modelos en las cuatro base de datos. }%
\begin{table}[ht!] \centering
  \scriptsize
  \begin{tabular}{c|cc|cc|cc|cc|c||c} 
 \multirow{2}{*}{$\hfrac{\text{\scriptsize Dataset}}{\text{Test size}}$} & \multicolumn{2}{c|}{Constant\es{e}$^\dagger$}& \multicolumn{2}{c|}{Elo$^\dagger$} & \multicolumn{2}{c|}{TrueSkill} & \multicolumn{2}{c|}{KickScore$^\dagger$} &  \multicolumn{2}{c}{TTT} \\
 & GM & $\log_2$BF & GM & $\log_2$BF & GM & $\log_2$BF & GM & $\log_2$BF & GM & LOOCV \\ \hline
\multirow{2}{*}{$\hfrac{\text{\scriptsize Ten\en{n}is}}{\num{186361}}$} & \multirow{2}{*}{$0.5593$} & \multirow{2}{*}{\num{7910}} & \multirow{2}{*}{$0.5695$} & \multirow{2}{*}{\num{3051}} & \multirow{2}{*}{$0.5722$} & \multirow{2}{*}{\num{1780}} & \multirow{2}{*}{$0.5758$} & \multirow{2}{*}{\num{93}} & \multirow{2}{*}{$\bm{0.5760}$} & \multirow{2}{*}{${0.5908}$} \\
 & & & & & & & & & & \\
 \multirow{2}{*}{$\hfrac{\text{\scriptsize Basketball}}{\num{20300}}$} & \multirow{2}{*}{$0.5006$} & \multirow{2}{*}{\num{1771}} & \multirow{2}{*}{$0.5305$} & \multirow{2}{*}{\num{72}} & \multirow{2}{*}{$0.5316$} & \multirow{2}{*}{\num{11}} & \multirow{2}{*}{$\bm{0.5328}$} & \multirow{2}{*}{-55} & \multirow{2}{*}{${0.5318}$} & \multirow{2}{*}{${0.5382}$} \\
  & & & & & & & & & & \\
 \multirow{2}{*}{$\hfrac{\text{\scriptsize Chess}}{\num{92004}}$} & \multirow{2}{*}{$0.3570$} & \multirow{2}{*}{\num{520}} & \multirow{2}{*}{$0.3552$} & \multirow{2}{*}{\num{1190}} & \multirow{2}{*}{$0.3580$} & \multirow{2}{*}{\num{148}} & \multirow{2}{*}{$\bm{0.3584}$} & \multirow{2}{*}{0} & \multirow{2}{*}{$\bm{0.3584}$} & \multirow{2}{*}{${0.3641}$} \\
 & & & & & & & & & \\
 \multirow{2}{*}{$\hfrac{\text{\scriptsize \en{Football}\es{F\'utbol}}}{\num{5759}}$} & \multirow{2}{*}{$0.3949$} & \multirow{2}{*}{\num{30}} & \multirow{2}{*}{$0.3867$} & \multirow{2}{*}{\num{204}} & \multirow{2}{*}{$0.3921$} & \multirow{2}{*}{\num{89}} & \multirow{2}{*}{$\bm{0.3961}$} & \multirow{2}{*}{\num{4}} & \multirow{2}{*}{$\bm{0.3963}$} &  \multirow{2}{*}{${0.3974}$} \\
  & & & & & & & & & & \\ \hline
  \end{tabular}
  \caption{
  \en{Model comparison in four databases. }%
  \es{Comparaciones de modelos en cuatro base de datos. }%
  \en{The GM columns report the geometric mean of the prior predictions of the models, Equation~\ref{eq:gmean}. }%
  \es{Las columnas GM reportan la media geométrica de las predicciones a priori de los modelos (ecuaci\'on \ref{eq:gmean}). }%
  \en{The $\log_2$BF columns report the Bayes Factor (Equation~\ref{eq:bayes_factor}) between TTT and the other models in the logarithmic scale. }%
  \es{Las columnas $\text{log}_2$BF reportan el Bayes Factor (ecuaci\'on \ref{eq:bayes_factor}) entre TTT y otros modelos en escala logarítmica. }%
  \en{For reference, the LOOCV column reports the geometric mean of the predictions using all historical data, $p(d_i| d_1, \dots, d_{i-1}, d_{i+1}, \dots, d_n, M)$. }%
  \es{Como referencia, la columna LOOCV reporta la media geométrica de las predicciones usando toda la informaci\'on hist\'orica, $p(d_i| d_1, \dots, d_{i-1}, d_{i+1}, \dots, d_n , M)$. }%

>>>>

% \newif\ifen
% \newif\ifes
% \newcommand{\en}[1]{\ifen#1\fi}
% \newcommand{\es}[1]{\ifes#1\fi}
% \entrue
\usepackage[utf8]{inputenc}
\usepackage{paracol}
\usepackage{multirow}

%% recommended packages orcidlink,
\usepackage{orcidlink,lmodern}

@@

% % by: https://tex.stackexchange.com/questions/74636/mla-package-and-thumbpdf
% \makeatletter
% \@namedef{ver@thumbpdf.sty}{}
% \makeatother

@@

\hypersetup{draft} %Only for diff checking
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks,
%     linkcolor={black!50!black},
%     citecolor={black!50!black},
%     urlcolor={black!80!black}
% }

@@

%% - \Abstract{} almost as usual
\Abstract{
%Humans develop complex skills through time.
Knowing how individual abilities change is essential in a wide range of activities. 
The most widely used skill estimators in industry and academia  (such as Elo and TrueSkill)  propagate information in only one direction, from the past to the future, preventing them from obtaining reliable initial estimates and ensuring comparability between estimates distant in time and space.
In contrast, the model TrueSkill Through Time (TTT) propagates all historical information throughout a single causal network, providing estimates with low uncertainty at any given time, enabling reliable initial skill estimates, and ensuring historical comparability.
% The improvement is evident since their prior predictions are several orders of magnitude more accurate than those obtained with other models.
% In addition to being intuitive for users, the TTT model achieves similar or even better results than more complex ones, such as KickScore.
Although the TTT model was published more than a decade ago, it was not available until now in the programming languages with the largest communities. 
Here we offer the first software for \proglang{Julia}, \proglang{Python}, and \proglang{R}, accompanied by a detailed overview for the general public and an in-depth scientific explanation. 
After illustrating its basic mode of use, we show how to estimate the learning curves of historical players of the Association of Tennis Professionals. 
Analytical approximation methods and message-passing algorithms allow inference to be solved efficiently using any low-end computer, even in causal networks with millions of nodes and irregular structures. 
} 
\Keywords{Learning, skill, Bayesian inference, gaming, education, sports, \proglang{Julia}, \proglang{Python},  \proglang{R}}

\Plainkeywords{Learning, skill, bayesian inference, gaming, education, sports, Julia, Python, R}

@@

\section{Introduction} \label{sec:intro}

Knowing how individual skills change over time is essential in the educational system and the labor market. 
Since skills are hidden variables, the best we can do is estimate them based on their direct observable consequences: the outcome of problem-solving and competitions. 
However, estimating learning curves is a sensitive issue, especially when they are used to make decisions that may impact individuals. 
Considering only the frequency of positive results as an indicator of the individual's ability could lead to wrong approximations, mainly because the outcome also depends on the difficulty of the challenge. 
For this reason, all widely used skill estimators use pairwise comparisons. 
With the first generative models proposed almost a century ago by~\cite{Thurstone1927} and~\cite{Zermelo1929}, it is assumed that the observed result probability $r$ depends on the performance $p$ of an agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$. 
The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, leading to a breakthrough that took place when~\cite{Elo2008} developed a methodology for the US Chess Federation (USCF), used by the International Chess Federation (FIDE) nowadays.

More recently, these models were extended and solved through the Bayesian approach to probability~\cite{glickman2001, Herbrich2007}.
%
Using Bayes's theorem we can quantify the uncertainty of the skill hypotheses using the information provided by the observed result and the specific model,
%
\begin{equation}\label{eq:event_inference}
 \underbrace{p(\overbrace{\text{Skill$_i$}}^{\text{Hidden}}|\overbrace{\text{Result}}^{\text{Observed}}, \text{Model})}_{\text{Posterior}} = \frac{\overbrace{P(\,\text{Result}\,|\,\text{Skill$_i$}\,,\text{Model})}^{\text{Likelihood}}\overbrace{p(\text{Skill$_i$})}^{\text{Prior}}}{\underbrace{P(\text{Result}\,|\,\text{Model})}_{\text{Evidence or prior prediction}}}
\end{equation}
%
where the only free variable is the skill hypothesis of agent $i$.
The prior quantifies the uncertainty about the skill before seeing the result, and the posterior quantifies the uncertainty after seeing the result.
The likelihood and the evidence can be interpreted as predictions of the observed results.
Because the evidence is the same for all hypotheses, the only factor that updates our beliefs is the likelihood.

\subsection{TrueSkill}

In this paper, we focus on a basic causal model in which skills generate the observable result of a model (Figure~\ref{fig:generative_model}).
The agents exhibit different performances at each event, varying around their actual skill, following a distribution $\N(p\,|\,s,\beta^2)$.
The model assumes that the agent with the highest performance wins, $r = (p_i > p_j)$.
%
\begin{figure}[ht!]
\centering \small
    \tikz{
    \node[det, fill=black!10] (r) {$r$} ;
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small Result:};
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

    \node[latent, above=of r, yshift=-0.45cm] (d) {$d$} ; %
    \node[const, right=of d] (dd) {\normalsize $ d = p_i-p_j$};
    \node[const, left=of d, xshift=-1.35cm] (d_name) {\small Difference:};

    \node[latent, above=of d, xshift=-0.8cm, yshift=-0.45cm] (p1) {$p_i$} ; %
    \node[latent, above=of d, xshift=0.8cm, yshift=-0.45cm] (p2) {$p_j$} ; %
    \node[const, left=of p1, xshift=-0.55cm] (p_name) {\small Performance:};

@@

    \node[const, left=of s1, xshift=-.85cm] (s_name) {\small Skill:};

@@

}
  \caption{Generative model in which skills cause the observable results mediated by the difference of hidden performances of both random variables around their unknown true skill ($d =p_i - p_j$).
  The one with the highest performance wins, $r = (d > 0)$. Observable variables are painted gray, hidden in white, and constants are shown as black dots. }
  \label{fig:generative_model}
\end{figure}
%
The parameter $\beta^2$, being the same for all agents, acts as the scale of the estimates: skills at a distance of one $\beta$ mean a \SI{76}{\percent} probability of winning, independent of the absolute value of the estimates.

For example, we consider a winning case ($p_i > p_j$) using a Gaussian prior (i.e.,~$\N(\,s\,|\,\mu, \sigma^2)$) for each skill.
Our prior belief about the difference in performances, $d=p_i-p_j$, is expressed as a Gaussian distribution centered on the difference in the prior estimates ($\mu_i -\mu_j$), with a variance that incorporates the uncertainty of both estimates ($\sigma_i$ and $\sigma_j$) and the variance of both performances ($\beta$), $\N( d \, | \, \mu_i -\mu_j \, ,\ 2\beta^2 + \sigma_i^2 + \sigma_j^2 \,)$. 
As we observed that the agent $i$ won, we know from the causal model that the hidden difference in performance was positive. 
Therefore, the prior prediction of the observed result, or evidence, is the cumulative density ($\Phi$) of all positive values of performance difference as expressed in Equation~\ref{eq:evidence}. 
During the rest of this work, the role of the model will be left implicit. 
%
\begin{equation}\label{eq:evidence}
 \overbrace{P(r)}^{\text{Evidence}} = 1-\Phi(0 \, | \overbrace{\mu_i^{\phantom{2}} - \mu_j}^{\hfrac{\text{Expected}}{\text{difference}}} , \, \overbrace{2\beta^2 + \sigma_i^2+ \sigma_j^2}^{\hfrac{\text{Total}}{\text{uncertainty}}})
\end{equation}
%
The evidence is a prediction made with all the prior hypotheses. 
Since it is a constant, the posterior uncertainty of each hypothesis is proportional to the product of their prior uncertainty and likelihood, as shown in Equation~\ref{eq:posterior_win}. 
Section~\ref{sec:2vs2} shows how these expressions are obtained by applying the sum and product rules. 
%
\begin{equation}\label{eq:posterior_win}
\underbrace{p(\,s_i\, | \, r \, )}_{\text{Posterior}} \propto \underbrace{1-\Phi(0 \, |  s_i - \mu_j , \, 2\beta^2 + \sigma_j^2)}_{\text{Likelihood} \ P(r|s_i)} \,  \underbrace{\N(s_i \, | \, \mu_i,\, \sigma_i^2)}_{\text{Prior} \ p(s_i)} 
\end{equation}
Dividing the right hand with the evidence $P(r)$ leads to the normalized posterior.
It is interesting to note the similarities and differences between likelihood and evidence. 
The likelihood quantifies the same cumulative density as the evidence but is centered on the difference between the hypothesis we are evaluating ($s_i$) and the opponent's mean estimate ($\mu_j$), with a variance that includes all uncertainties except the one of $s_i$. 
%
\begin{figure}[ht!]
    \centering
    \includegraphics[page={1},width=.6\linewidth]{figures/posterior_win}
    \caption{Update of the belief for the winning case. 
    The proportional posterior is the product of the prior (Gaussian) and the likelihood (cumulative Gaussian). 
    The evidence is the integral of the proportional posterior. 
    The distributions are not necessarily on the same scale: the prior integrates to $1$, while the likelihood goes from $0$ to $1$. 
    }
    \label{fig:posterior_win}
\end{figure}

The posterior is just the prior's density that is not filtered by the likelihood. 
The surprise, defined as the likelihood's complement, works as a filter for the prior. 
In the region of very high-skill hypotheses, where the winning result would have generated almost no surprise ($\lim_{s_i \to \infty}P(r|s_i) = 1$), the posterior receives all the prior's density. 
In the region of low-skill hypotheses, a win would generate a great surprise ($\lim_{s_i \to -\infty}P(r|s_i) = 0$), and the posterior receives no density from the prior.

It is important to stress that the posterior, although similar, is not a Gaussian distribution, preventing us from using Equation~\ref{eq:posterior_win} iteratively. 
But due to the shape of the exact posterior, a Gaussian distribution could be used as a good approximation, allowing us to avoid the computational cost of the sampling methodologies. 
The success of the TrueSkill solution~\citep{Herbrich2007} is based on the usage of an efficient method for computing the Gaussian distribution that best approximates the exact posterior~(see Section~\ref{sec:approximate_posterior}),
\begin{equation} \label{eq:approx} 
\widehat{p}(s_i| r, s_j) = \underset{\mu, \sigma}{\text{ arg min }} \ \ \text{KL}(\, p(s_i| r, s_j) \, || \,  \N(s_i|\mu, \sigma^2) \, )
\end{equation}
%
in terms of Kullback-Leibler divergence minimization between the true and the approximate distribution. 
This method allows us to efficiently apply Equation~\ref{eq:posterior_win} iteratively over a sequence of observations, which would otherwise be infeasible. 
The approach adopted by TrueSkill to treat the dynamical process, known as \emph{filtering}, uses the last approximate posterior as the prior for the next event. 
The approximate posterior at any given time is defined as:
%
\begin{equation}\label{eq:filter} %\tag{\text{filtering}}
 \widehat{\text{Posterior}}_t \propto \widehat{\text{Likelihood}}_t  \overbrace{\widehat{\text{Likelihood}}_{t-1} \dots \underbrace{\widehat{\text{Likelihood}}_{1} \text{Prior}_1}_{\widehat{\text{Posterior}}_{1} \text{ as } \text{Prior}_{2}} }^{\widehat{\text{Posterior}}_{t-1} \text{ as } \text{Prior}_{t}} %= \text{Prior}_1 \prod_{i=1}^t \text{Likelihood}_i 
\end{equation}
%
where {\footnotesize $\widehat{\text{Posterior}}_i$} and {\footnotesize $\widehat{\text{Likelihood}}_i$} represent the approximations induced by the Equation~\ref{eq:approx} at the $i$-th event. 
If we consider the likelihood as a filter of the prior, each posterior is the accumulation of all previous filters. 
In this way, information propagates from past to future estimates. 
Since skills change over time, it is necessary to incorporate some uncertainty $\gamma$ after each step. 
%
\begin{equation}\label{eq:dynamic_factor}
 \widehat{p}(s_{i_t}) = \N(s_{i_t} | \mu_{i_{t-1}}, \sigma_{i_{t-1}}^2 + \gamma^2 )
\end{equation}
Because the filtering approach is an ad-hoc procedure that does not arise from any probabilistic model, its estimates have some problems. 
The most obvious is that the beginning of any sequence of estimates always has high uncertainty. 
But temporal and spatial decouplings may also occur, preventing the comparison between distant estimates. 
Although the relative differences between current estimates within well-connected communities are correct, estimates separated in time and between poorly connected communities may be incorrect. 
All of these issues are related to the fact that information propagates in only one direction through the system and can be solved by inferring with the available information from events occurring in parallel jointly with future events.
